\MakeUppercase{The Structure of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-07-04 k-planar.tex
Working....

\begin{thm}\thmlabel{generalized-tripod}
  There exists a tree $T$ rooted at a node $x_0$ where each node $x\in V(T)$ is associated with a subset $Y_x\subseteq V(G)$ that defines several related sets $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$ that satisfy the following properties:
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $t\ell$.
    \item\ylabel{separator} There is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} $x$ has $t'\le t$ $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{thm}

Before proving \thmref{generalized-tripod} we point out another properties that are immediately implied by it:

\begin{compactenum}[(Y1)]\setcounter{enumi}{3}
  \item\ylabel{containment} $V_x\subseteq V_a$ and $N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
\end{compactenum}

@@

@@

  Let $(Q_x: x\in V(I))$ denote the given $I$-partition of $G$.  For a node $x\in V(T)$, let $R_x:=\bigcup_{y\in C_x} Q_{y}$. For a node $x\in V(T)$, we say that a $(k,d)$ $vw$-shortcut $P$ in $G$ \emph{touches} $x$ if $V(P)\cap R_x\neq\emptyset$.  If $G$ contains a $vw$-shortcut that touches $x$, then we say that $v$ and $w$ \emph{participate} in $x$.

>>>>

\begin{thm}\thmlabel{generalized-tripod}
  There exists a tree $T$ rooted at a node $x_0$ where each node $x\in V(T)$ is associated with a subset $Y_x\subseteq V(G)$ that defines several related sets $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$ that satisfy the following properties:
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $t\ell$.
    \item\ylabel{separator} There is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} $x$ has $t'\le t$ strict $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{thm}

Before proving \thmref{generalized-tripod} we point two another properties that are immediately implied by it:

\begin{compactenum}[(Y1)]\setcounter{enumi}{3}
  \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$. 
  \item\ylabel{containment} $V_x\subseteq V_a$ and $N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
\end{compactenum}

@@

Let $T$ and, for each $x\in V(T)$, $Y_x$, $V_x$, $F_x$, and $N_x$ be the decomposition obtained by applying \thmref{generalized-tripod} ot $G$.
For a node $x\in V(T)$, we say that a $(k,d)$ $vw$-shortcut $P$ in $G$ \emph{touches} $x$ if $v\in Y_x$ or $V_x$ contains an internal vertex of $P$.  If $G$ contains a $vw$-shortcut that touches $x$, then we say that $v$ and $w$ \emph{participate} in $x$.

For any $v\in V(G)$, the union of all $(k,d)$ shortcuts in $G$ that have $v$ as an endpoint is a connected subgraph of $G$.  It follows that the set of nodes $x\in V(T)$ in which $v$ participates is a non-empty (connected) subtree of $T$ and we let $a(v)$ denote the root of this subtree. For each $x\in V(T)$, we define the \emph{separator}
\[  
     S_x := \{v\in V(G): a(v)= x\} \enspace .
\]
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $I:=G^{k,d}/(S_x : x\in V(T))$ denote the resulting quotient graph and we let $V(I)\subseteq V(T)$ in the obvious way, so that each $x\in V(I)$ is the vertex obtained by contracting $S_x$ in $G^{k,d}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $I$.)

\begin{clm}
  For every $x\in V(T)$, $S_x\subseteq N_x$.
\end{clm}

\begin{proof}
  For the sake of contradiction, assume otherwise.
  Then, $G$ contains a $(k,d)$ $vw$-shortcut $P$ that contains an internal vertex $v'\in Y_x$ and such that $w\not\in N_x$.  By \yref{separator}, $P$ contains an internal vertex $v''\in F_x$.  By \yref{ancestor-edge}, $v''\in Y_a$ for some strict $T$-ancestor $a$ of $x$, contradicting the assumption that $v\in S_x$ (since $v$ would would be in $Y_{a'}$ for some $T$-ancestor $a'$ of $a$). 
\end{proof}

Next we show that $\mathcal{S}$ has small layered width with respect to $\mathcal{L}$:

\begin{clm}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(I)$, $|S_x\cap V_i|\le t\ell(1+(d-1)^{k-1})$.  
\end{clm}

\begin{proof}
  Consider some vertex $v\in S_x\cap L_i$.  There are two cases to consider:
  \begin{enumerate}
    \item $v\in Y_x$.  By \yref{thickness}, there are at most $t\ell$ such vertices in $S_x\cap L_i$.
    \item $G$ contains a $(k,d)$ $vw$-shortcut $P$ and $Y_x$ contains an internal vertex $v'$ of $P$. In this case, $v'\in L_{i+j}\cap Y_x$ for some $j\in\{-k+1,\ldots,k-1\}$.   Since $|L_{i+j}\cap Y_x|\le t\ell$, the length of $P$ is at most $k-1$, and the internal vertices of $P$ have degree at most $d$ in $G$, there are at most $t\ell(d-1)^{k-1}$ such vertices in $S_x\cap L_i$. \qedhere
  \end{enumerate}
\end{proof}

\begin{clm}
  For each edge $xy\in E(I)$, one of $x$ or $y$ is a $T$-ancestor of the other.
\end{clm}

\begin{proof}
  Consider the lowest common $T$-ancestor $a$ of $x$ and $y$ and assume, for the sake of contradiction that $a\not\in\{x,y\}$.

  By \yref{separator}, $G-(F_a\cup Y_a)$ does not contain any path from a vertex $v\in V_x$ to a vertex $w\in V_y$.  
\end{proof}

\begin{clm}
  The graph $I$ has a tree-decomposition whose largest bag has size at most $t^k$.
\end{clm}

\begin{proof}
  For the tree-decomposition of $I$ we use the same tree $T$ used in the tree-decomposition of $H$.  First we claim that, for each edge

\end{proof}

\newpage

@@

  Let $(Q_x: x\in V(I))$ denote the given $I$-partition of $G$.  For a node $x\in V(T)$, let $R_x:=\bigcup_{y\in C_x} Q_{y}$.

  For a node $x\in V(T)$, we say that a $(k,d)$ $vw$-shortcut $P$ in $G$ \emph{touches} $x$ if $V(P)\cap R_x\neq\emptyset$.  If $G$ contains a $vw$-shortcut that touches $x$, then we say that $v$ and $w$ \emph{participate} in $x$.
